Measurement Bias from Address Aliasing

### 2014-08-15 paper.tex
Clarify explanation on execution port vacancy
Hopefully easier to follow, comment from one of the reviewers that
the previous explanation was unclear.

While optimizations such as these are good on the average case, there are corner cases. 
In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096 bytes.
A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, causing the load to be reissued.
The number of times this happens can be counted by the following event:
\begin{describe}{{pad}}
  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS.}] \emph{``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.''} \cite[B.3.4.4]{OptimizationManual}
\end{describe}
This event is listed in the manual for microarchitectures going back to Nehalem i7, including Ivy Bridge and Haswell CPUs~\cite{Volume3B}.
Older architectures such as Core 2 does not have this particular event listed, however 4K aliasing is covered by a more general event:
\begin{describe}{{pad}}
  \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''} \cite[Table 19-17]{Volume3B}
\end{describe}

@@

We see the most extreme change from median to worst case in the number of alias events. 
If we plot the graph for address aliasing, we see that it is near zero everywhere and spikes at exactly the points we observe bias. 
The results show a high number of resource stalls and pending memory loads when the spikes occur, which is consistent with address aliasing issues.
On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from around 272,000 to 136,000.
The reservation station buffers micro-ops for scheduling to the execution units, and a stall event means that there are no free slots available~\cite[Table 19-2]{Volume3B}.
Fewer stalls in the aliasing case could indicate less contention on the reservation station.
This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the in total 8 execution ports have fewer operations going in parallel.
Note that the number of micro-ops \emph{retired} overall does not change. %, but for some reason the micro-ops stay in the execution pipeline for a shorter time.
Bacause the memory disambiguation system does not handle address aliasing, the potential for out of order execution is limited~\cite[Page 2-20]{OptimizationManual}.
We assume that the tiny kernel size immediately prevents instructions from executing out of order, leading to the lower counts.
%Our assumption is that this is caused by the CPU being more conservative in the aliasing case, issuing fewer operations (speculatively) in parallel.
%With fewer instructions per cycle, there is also less pressure on the reservation station.

>>>>

While optimizations such as these are good on the average case, there are corner cases. 
In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096 bytes.
A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, causing the load to be reissued.
The number of times this happens can be counted by the following event:
\begin{describe}{{pad}}
  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS.}] \emph{``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.''} \cite[B.3.4.4]{OptimizationManual}
\end{describe}
This event is listed in the manual for microarchitectures going back to Nehalem i7, including Ivy Bridge and Haswell CPUs~\cite{Volume3B}.
Older architectures such as Core 2 do not have this particular event listed, however 4K aliasing is covered by a more general event:
\begin{describe}{{pad}}
  \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''} \cite[Table 19-17]{Volume3B}
\end{describe}

@@

We see the most extreme change from median to worst case in the number of alias events. 
If we plot the graph for address aliasing, we see that it is near zero everywhere and spikes at exactly the points we observe bias. 
The results show a high number of resource stalls and pending memory loads when the spikes occur, which is consistent with address aliasing issues.
On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from around 272,000 to 136,000.
The reservation station buffers micro-ops for scheduling to the execution units, and a stall event means that there are no free slots available~\cite[Table 19-2]{Volume3B}.
Fewer stalls in the aliasing case could indicate less contention on the reservation station.
This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the in total 8 execution ports have fewer operations going in parallel.
Note that the number of micro-ops \emph{retired} overall does not change.
The memory disambiguation system automatically assumes dependency between a load and a previous store that are aliased, which can limit the potential for issuing many instructions simultaneously and out of order~\cite[Page 2-20]{OptimizationManual}.
In our case, lower occupation of execution ports and less pressure on the reservation station can be explained by the CPU not being able to issue more operations while waiting for an aliased memory access to be resolved.
